The Effects of Labor-Associated Drug Interventions on Neonatal Health Outcomes: A Predictive Analysis of Apgar5 Scores

### 2025-02-17 infant_health/paper/paper.qmd

---
title: "The Effects of Labor-Associated Drug Interventions on Neonatal Health Outcomes: A Predictive Analysis of Apgar5 Scores"
subtitle: "Steroids Enhance Neonatal Health, While Antibiotics Show Limited Effectiveness"
author: 
  - Yunkyung Ko
thanks: "Code and data are available at: [https://github.com/koyunkyung/infant_health](https://github.com/koyunkyung/infant_health)."
date: today
date-format: long
abstract: "How maternal and neonatal drug interventions during labor shape the health status of a newborn remains a key question in perinatal care. With the 2023 Natality Data for the United States, the relationship between six labor-related treatments and Apgar5 scores, a standard measure of neonatal health, was examined. Using a Random Forest model for prediction and a Bayesian Linear Model for inference, we identified usage of steroids and chorioamnionitis as the most impactful, while antibiotics dosage showed minimal influence. The analysis shows disparities in treatment efficacy, and further suggests the need for optimized corticosteroid formulations or targeted antibiotic protocols."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
toc-title: "Table of Contents"
toc-depth: 2
toc-location: left
---

@@

By structuring these variables into categorical or numerical forms, the data can be analyzed and used to model the relationship between treatments and the APGAR5 score. Specifically, drug usage during labor and delivery was encoded as binary variables (@sec-predictor). However, this has a limitation that it disregards the quantitative intensity, dosage, or frequency of drug administration, leading to incomplete modeling of the treatment's impact [@geeksforgeeks_feature_encoding]. This simplification may also amplify noise in the data by grouping dissimilar observations into the same category, reducing the model's ability to detect subtle relationships [@geeksforgeeks_feature_encoding].

@@

@fig-apgar1 shows that the majority of infants in the raw dataset achieve a high APGAR5 score, clustering around 9 and 10. Very few observations exist for lower scores, reflecting rare instances of significant distress at birth. Even after filtering the dataset by selecting the relevant variables for analysis and removing the NA values, @fig-apgar2 shows that the observations are overly clustered around high APGAR5 scores.

@@

All of the predictor variables used in the analysis are classified under the same category, which is the 'Characteristics of labor and delivery' [@natality2023]. This item, which contains 6 separate checkboxes that the respondent can choose from, allows for the reporting of more than one characteristic and includes a choice of "None of the above" [@cdcFacilityGuide].

@@

@fig-numtreat shows that the majority of observations involve 0 to 2 treatments during labor and delivery, with a steep decline in counts for 3 or more treatments. Most births reported in this dataset occur with minimal medical intervention, and higher number of treatments are relatively rare.

@@

@fig-typetreat shows that the distribution of treatment types varies significantly, with anesthesia being the most commonly recorded intervention, while chorioamnionitis is rare. Treatments like anesthesia and antibiotics, with higher observation counts, may have a greater impact on the observed outcomes, whereas less frequent interventions, such as steroids or chorioamnionitis, may require careful consideration to avoid biases due to smaller sample sizes [@selectstatistics2024].

@@

### Clinical chorioamnionitis diagnosed during labor or maternal temperature over 38 degrees celcius (100.4 degrees fahrenheit)

@@

```{r}
#| label: fig-corr1
#| fig-cap: Mosaic plot showing the correlation between induction and augmentation of labor (**Note:** the shading intensity (from light green to dark green) represents the standardized residuals of the chi-squared test for the contingency table.)
#| echo: false
#| warning: false
#| message: false

@@

```{r}
#| label: fig-corr2
#| fig-cap: Mosaic plot showing the correlation between usage of steroids and antibiotics (**Note:** the shading intensity (from light green to dark green) represents the standardized residuals of the chi-squared test for the contingency table.)
#| echo: false
#| warning: false
#| message: false

@@

In @fig-corr3, the large, darkly shaded area for "Chorioamnionitis:0" and "Antibiotics:0" shows that cases where neither condition is present are more common than expected [@friendly1992]. Conversely, the smaller and lighter areas, especially for "Chorioamnionitis:1" and "Antibiotics:1", suggest that when chorioamnionitis occurs, antibiotics are often used. The pattern reflects a likely positive correlation [@friendly1992] between chorioamnionitis and anitbiotics usage.

```{r}
#| label: fig-corr3
#| fig-cap: Mosaic plot showing the correlation between usage of chorioamnionitis and antibiotics (**Note:** the shading intensity (from light green to dark green) represents the standardized residuals of the chi-squared test for the contingency table.)
#| echo: false
#| warning: false
#| message: false

@@

The Random Forest model is a non-parametic method that predicts $y_i$ using a collection of decision trees [@mlnuggets2024]. The predicted 5-minute Apgar score is modeled as:

@@

The model is implemented using the `stan_glm` function in the `rstanarm` package [@rstanarm]. Markov Chian Monte Carlo (MCMC) sampling was conducted with 4 chains and 2000 iterations per chain. Convergence diagnostics were verified, and posterior predictive checks were conducted to assess model fit, as shown in @sec-model-details.

@@

```{r}
#| label: fig-result2
#| fig-cap: Heatmap of the overall correlation between treatments administered during labor/delivery and Apgar5 scores
#| echo: false
#| warning: false
#| message: false

@@

@fig-chorscore shows that the observation count peaks sharply at a score of 1, then stabilizes between scores 2 and 5, followed by a gradual decline from scores 6 through 10. This distribution suggests that infants that were diagnosed clinical chorioamnionitis during labor or maternal temperature over 38 degrees celcius tend to achieve lower-to-mid range Apgar5 scores.

@@

@tbl-rfsummary1 and @fig-rfsumbar shows that usage of steroids (45.76%) is the most influential factor among the treatments during labor and delivery when predicting Apgar5 scores. It is followed by usage of chorioamnionitis (22.44%) and anesthesia (13.99%), which also contributes substantially. Meanwhile, the exposure to antibiotics (0.00%) shows no measurable impact in this context, and suggests its irrelevance to the model's predictions.

@@

```{r}
#| echo: false
#| eval: true
#| label: fig-rfsumbar
#| fig-cap: Horizontal bar plot showing the variable importance of treatments during labor and delivery on infant health
#| warning: false

@@

## Optimization of anetenatal corticosteroid formulations {#sec-first-point}

@sec-randfor-result reveals that maternal and neonatal treatments during labor, such as steroids and antibiotics, have differential impacts on infant health outcomes. Usage of steroids, which showed positive correlations with the Apgar5 score in @fig-sterscore, was actually affirmed their critical role in improving neonatal survival, especially in extremely preterm infants [@nih_prenatal_steroid_treatment]. In particular, antenatal corticosteroids stand out as the core intervention for preterm births under the gestation period of 22 weeks, significantly reducing complications like respiratory distress syndrome by accelerating fetal lung development [@nih_prenatal_steroid_treatment].

In accordance to these analyses results, pharmaceutical companies can focus on optimizing steroid formulations for safer and more effective administration during earlier gestations [@nih_prenatal_steroid_treatment]. Current formulations of antenatal corticosteroids, like betamethasone, have shown strong efficacy in improving neonatal outcomes but may carry risks when administered too early or in repeated doses [@nichd_prenatal_steroid_treatment]. Hence, adjusting dosages or delivery mechanisms could enhance safety while preserving their life-saving benefits for preterm infants. For instance, this could involve sustained-release versions or targeted delivery systems to minimize potential side effects on both the mother and fetus [@nichd_prenatal_steroid_treatment].

@@

In order to accurately contextualize the findings of this analysis, several limitations must be acknowledged. First, the observational nature of the study may introduce confounding variables that were not accounted for, potentially biasing the results [@biostatistics_confounding]. Unlike randomized controlled trials (RCTs), observational studies lack random assignment, making it challenging to establish causality [@pmc5073487]. Confounding factors, such as socioeconomic status, pre-existing health conditions, or access to healthcare, could influence both the treatment received and the neonatal outcomes, thereby skewing the correlations observed [@biostatistics_confounding].

@@

The 2023 Natality Dataset [@rawdata] provides detailed records of maternal and neonatal health outcomes across diverse populations in the United States. This obervational dataset aggregates data from hospital records, birth certificates, and surveys completed by healthcare providers [@natality2023]. Its comprehensive scope includes variables like maternal age, gestational age, socio-economic status, and the application of specific labor-related treatments [@cdc_facility_worksheet_2016].

@@

The 'Infant Health Prediction API' provides predictions for infant health status based on maternal and delivery-related factors. By accepting structured datasets in CSV format, it offers researchers a powerful tool to generate real-time predictions using advanced maching learning models.

>>>>

---
title: "The Effects of Labor-Associated Drug Interventions on Neonatal Health Outcomes: A Predictive Analysis of Apgar5 Scores"
subtitle: "Steroids enhance neonatal health, while antibiotics show limited effectiveness"
author: 
  - Yunkyung Ko
thanks: "Code and data are available at: [https://github.com/koyunkyung/infant_health](https://github.com/koyunkyung/infant_health)."
date: today
date-format: long
abstract: "How maternal and neonatal drug interventions during labor shape the health status of a newborn remains a key question in perinatal care. With the 2023 Natality Data for the United States, the relationship between six labor-related treatments and Apgar5 scores, a standard measure of neonatal health, was examined. Using a Random Forest model for prediction and a Bayesian Linear Model for inference, we identified usage of steroids and chorioamnionitis as the most impactful, while antibiotics dosage showed minimal influence. The analysis shows disparities in treatment efficacy, and further suggests the need for optimized corticosteroid formulations or targeted antibiotic protocols."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
toc-title: "Table of Contents"
toc-depth: 2
toc-location: left
---

@@

By structuring these variables into categorical or numerical forms, the data can be analyzed and used to model the relationship between treatments and the Apgar5 score. Specifically, drug usage during labor and delivery was encoded as binary variables (@sec-predictor). However, this has a limitation that it disregards the quantitative intensity, dosage, or frequency of drug administration, leading to incomplete modeling of the treatment's impact [@geeksforgeeks_feature_encoding]. This simplification may also amplify noise in the data by grouping dissimilar observations into the same category, reducing the model's ability to detect subtle relationships [@geeksforgeeks_feature_encoding].

@@

@fig-apgar1 shows that many infants in the raw dataset achieve a high Apgar5 score, clustering around 9 and 10. Very few observations exist for lower scores, reflecting rare instances of significant distress at birth. Even after filtering the dataset by selecting the relevant variables for analysis and removing the NA values, @fig-apgar2 shows that the observations are overly clustered around high Apgar5 scores.

@@

All the predictor variables used in the analysis are classified under the same category, which is the 'Characteristics of labor and delivery' [@natality2023]. This item, which contains 6 separate checkboxes that the respondent can choose from, allows for the reporting of more than one characteristic and includes a choice of "None of the above" [@cdcFacilityGuide].

@@

@fig-numtreat shows that many observations involve 0 to 2 treatments during labor and delivery, with a steep decline in counts for 3 or more treatments. Most births reported in this dataset occur with minimal medical intervention, and higher number of treatments are relatively rare.

@@

@fig-typetreat shows that the distribution of treatment types varies significantly, with anesthesia being the most recorded intervention, while chorioamnionitis is rare. Treatments like anesthesia and antibiotics, with higher observation counts, may have a greater impact on the observed outcomes, whereas less frequent interventions, such as steroids or chorioamnionitis, may require careful consideration to avoid biases due to smaller sample sizes [@selectstatistics2024].

@@

### Clinical chorioamnionitis diagnosed during labor or maternal temperature over 38 degrees Celcius (100.4 degrees Fahrenheit)

@@

```{r}
#| label: fig-corr1
#| fig-cap: The correlation between induction and augmentation of labor (**Note:** the shading intensity (from light green to dark green) represents the standardized residuals of the chi-squared test for the contingency table.)
#| echo: false
#| warning: false
#| message: false

@@

```{r}
#| label: fig-corr2
#| fig-cap: The correlation between usage of steroids and antibiotics (**Note:** the shading intensity (from light green to dark green) represents the standardized residuals of the chi-squared test for the contingency table.)
#| echo: false
#| warning: false
#| message: false

@@

In @fig-corr3, the large, darkly shaded area for "Chorioamnionitis:0" and "Antibiotics:0" shows that cases where neither condition is present are more common than expected [@friendly1992]. Conversely, the smaller and lighter areas, especially for "Chorioamnionitis:1" and "Antibiotics:1", suggest that when chorioamnionitis occurs, antibiotics are often used. The pattern reflects a likely positive correlation [@friendly1992] between chorioamnionitis and antibiotics usage.

```{r}
#| label: fig-corr3
#| fig-cap: The correlation between usage of chorioamnionitis and antibiotics (**Note:** the shading intensity (from light green to dark green) represents the standardized residuals of the chi-squared test for the contingency table.)
#| echo: false
#| warning: false
#| message: false

@@

The Random Forest model is a non-parametric method that predicts $y_i$ using a collection of decision trees [@mlnuggets2024]. The predicted 5-minute Apgar score is modeled as:

@@

The model is implemented using the `stan_glm` function in the `rstanarm` package [@rstanarm]. Markov Chain Monte Carlo (MCMC) sampling was conducted with 4 chains and 2000 iterations per chain. Convergence diagnostics were verified, and posterior predictive checks were conducted to assess model fit, as shown in @sec-model-details.

@@

```{r}
#| label: fig-result2
#| fig-cap: The overall correlation between treatments administered during labor/delivery and Apgar5 scores
#| echo: false
#| warning: false
#| message: false

@@

@fig-chorscore shows that the observation count peaks sharply at a score of 1, then stabilizes between scores 2 and 5, followed by a gradual decline from scores 6 through 10. This distribution suggests that infants that were diagnosed clinical chorioamnionitis during labor or maternal temperature over 38 degrees Celcius tend to achieve lower-to-mid range Apgar5 scores.

@@

@tbl-rfsummary1 and @fig-rfsumbar shows that usage of steroids (45.76%) is the most influential factor among the treatments during labor and delivery when predicting Apgar5 scores. It is followed by usage of chorioamnionitis (22.44%) and anesthesia (13.99%), which also contributes substantially. Meanwhile, the exposure to antibiotics (0.00%) shows no measurable impact and suggests its irrelevance to the model's predictions.

@@

```{r}
#| echo: false
#| eval: true
#| label: fig-rfsumbar
#| fig-cap: Variable importance of treatments during labor and delivery on infant health, expressed in horizontal bar plots
#| warning: false

@@

## Optimization of antenatal corticosteroid formulations {#sec-first-point}

@sec-randfor-result reveals that maternal and neonatal treatments during labor, such as steroids and antibiotics, have differential impacts on infant health outcomes. Usage of steroids, which showed positive correlations with the Apgar5 score in @fig-sterscore, was affirmed their critical role in improving neonatal survival, especially in extremely preterm infants [@nih_prenatal_steroid_treatment]. In particular, antenatal corticosteroids stand out as the core intervention for preterm births under the gestation period of 22 weeks, significantly reducing complications like respiratory distress syndrome by accelerating fetal lung development [@nih_prenatal_steroid_treatment].

In accordance with these analyses results, pharmaceutical companies can focus on optimizing steroid formulations for safer and more effective administration during earlier gestations [@nih_prenatal_steroid_treatment]. Current formulations of antenatal corticosteroids, like betamethasone, have shown strong efficacy in improving neonatal outcomes but may carry risks when administered too early or in repeated doses [@nichd_prenatal_steroid_treatment]. Hence, adjusting dosages or delivery mechanisms could enhance safety while preserving their life-saving benefits for preterm infants. For instance, this could involve sustained-release versions or targeted delivery systems to minimize potential side effects on both the mother and fetus [@nichd_prenatal_steroid_treatment].

@@

To accurately contextualize the findings of this analysis, several limitations must be acknowledged. First, the observational nature of the study may introduce confounding variables that were not accounted for, potentially biasing the results [@biostatistics_confounding]. Unlike randomized controlled trials (RCTs), observational studies lack random assignment, making it challenging to establish causality [@pmc5073487]. Confounding factors, such as socioeconomic status, pre-existing health conditions, or access to healthcare, could influence both the treatment received and the neonatal outcomes, thereby skewing the correlations observed [@biostatistics_confounding].

@@

The 2023 Natality Dataset [@rawdata] provides detailed records of maternal and neonatal health outcomes across diverse populations in the United States. This observational dataset aggregates data from hospital records, birth certificates, and surveys completed by healthcare providers [@natality2023]. Its comprehensive scope includes variables like maternal age, gestational age, socio-economic status, and the application of specific labor-related treatments [@cdc_facility_worksheet_2016].

@@

The 'Infant Health Prediction API' provides predictions for infant health status based on maternal and delivery-related factors. By accepting structured datasets in CSV format, it offers researchers a powerful tool to generate real-time predictions using advanced matching learning models.
